Fail Faster

### 2025-03-01 paper/paper.tex
changes

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigplan,screen]{acmart}%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

Our solution is \name, a generator library that preserves
\bq’s functionality while improving performance through
\textit{staged metaprogramming} (staging). Staging is a lightweight,
domain-specific compilation technique that allows us to completely eliminate the
runtime overhead of many common generator abstractions. For instance, in
\name, monadic operations are zero-cost.
\name’s staged eliminates unnecessary allocations,
generates static control structures, and leverages unboxed integer libraries to
achieve performance improvements over \texttt{base\_quickcheck}, making it
well-suited to real-world PBT applications.

@@

\section{Background}
\subsection{Staged Metaprogramming}
\subsection{Property-Based Testing}
\subsection{Existing PBT Generators}

\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq\ is the best tool to reproduce.}

@@

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = size:int -> random:SR.t -> 'a}. Size and random seed to deterministic value.
      \item This gets a monad intance in the obvious way
    \end{itemize}
  \end{itemize}
}

@@

>>>>

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:

%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigplan,screen]{acmart}%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

Our solution is \name, a generator library that preserves
\bq’s functionality while improving performance through
\textit{staged metaprogramming} (staging). Staging is a lightweight,
domain-specific compilation technique that allows us to completely eliminate the
runtime overhead of many common generator abstractions. \jwc{This relies on the fact that the code of generators is known statically at compile time} For instance, in
\name, monadic operations are zero-cost.
\name’s staged eliminates unnecessary allocations,
generates static control structures, and leverages unboxed integer libraries to
achieve performance improvements over \texttt{base\_quickcheck}, making it
well-suited to real-world PBT applications.

@@

\section{Background}
\subsection{Staged Metaprogramming}

\jwc{A good quick presentation of staged metaprogramming is here, in section 6.
https://www.cl.cam.ac.uk/~jdy22/papers/a-typed-algebraic-approach-to-parsing.pdf
}

\subsection{Property-Based Testing}
\subsection{Other Languages and Libraries}

\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq{} is the best tool to reproduce.}

\jwc{Note that this explanation require some prior note of exactly *why* BQ is slow... i.e. the abstraction overhead of the library is high.}

\jwc{
  \begin{itemize}
    \item Scala. Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS. Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
    % \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
    % GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
    % the highest-performance generators.
    % \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
  \end{itemize}
}

@@

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = size:int -> random:SR.t -> 'a}. Size and random seed to deterministic value.
      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
      \item This gets a monad intance in the obvious way (show code).
    \end{itemize}
    \item Note that \emph{extensionally} \texttt{generate (create (fun ~size ~random -> e)) ~size ~random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
  \end{itemize}
}

In OCaml, garbage collection is the killer.

This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.

\begin{lstlisting}
let gen_list_bool =
  fixed_point (fun gl ->
    let%bind n = size in
    if n <= 0 then return [] else
      let%bind x = bool in
      let%bind xs = with_size ~size:(n-1) gl in
      return (x :: xs)
  )
\end{lstlisting}

On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
\begin{lstlisting}
  let rec gen_list_bool_faster ~size ~random  = 
    if size <= 0 then []
    else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
\end{lstlisting}

@@

\jwc{
  \begin{itemize}
    \item To eliminate this runtime overhead and generate efficient code, we do a usual ``binding-time analysis''.
    \item The crux of this binding-time analysis is that while generators are `just' values of type \texttt{'a t}, in practice they are always \emph{completely statically defined}: one never
    computes a generator dynamically at test time.
  \end{itemize}
}

\jwc{Here!}

\begin{lstlisting}
type 'a gen = int code -> Random.t code -> 'a
\end{lstlisting}
